The Timing and Likelihood of the Origin of Life from Earth's Impact History

### 2023-04-26 ms.tex
fixed end of discussion/conclusions section

As illustrated in Figure \ref{fig:example_impact_histories}, a life-starting impact does not occur in every simulated impact history. In our impact-driven model for the origin of life, biopoesis requires (1) an impact of sufficient mass to make prebiotic nitriles and (2) the lack of a later ocean-vaporizing collision that destroys the biosphere without rekindling it. Figure \ref{fig:probabilities_of_impacts} shows the probability of conditions (1) and (2) occurring as a function of the minimum impact mass that produces prebiotic molecules. For example, consider the \citet{Wogan_2023} optimistic minimum mass of $4 \times 10^{20}$ kg indicated in Figure \ref{fig:probabilities_of_impacts}. All simulated impact histories have at least one impact bigger than $4 \times 10^{20}$ kg (blue line in Figure \ref{fig:probabilities_of_impacts}). However, the probability of the origin of life in this scenario is 92\%, because 8\% of the time the last $4 \times 10^{20}$ kg collision is followed by a smaller ocean-vaporizing impact that perhaps sterilizes the planet (orange dashed line in Figure \ref{fig:probabilities_of_impacts}). An origin of life is less probable if we consider the \citet{Wogan_2023} pessimistic minimum mass ($> 5 \times 10^{21}$ kg) to produce significant origin of life nitriles. A $> 5 \times 10^{21}$ kg impact without later ocean vaporization occurs 41\% of the time. A life-starting impact was therefore not an anomalous event on the early Earth, and was strongly favored (92\% probability) under optimistic assumptions.

@@

% The timing of the last life-starting impact was not very sensitive to $f_\mathrm{E,vap}$ (Appendix Figure \ref{fig:timing_sensitivity}). Smaller $f_\mathrm{E,vap}$ favored a life-starting impactor earlier in the Hadean because small $f_\mathrm{E,vap}$ reduces the chance of late ocean-vaporization. Regardless, the median time of impact is always between 4.29 and 4.41 Ga which 95\% uncertainties spanning the Hadean eon.

% The probability of an impact that produces prebiotic molecules without later ocean vaporization (i.e., a life-starting impact) is somewhat sensitive to $f_\mathrm{E,vap}$ (Appendix Figure \ref{fig:probabilities_of_impacts_sens}). Smaller $f_\mathrm{E,vap}$ makes ocean-vaporization more rare which increases the chance of a life-starting impact.

@@

Our simulations of Earth's impact history do not always result in an origin of life. There are some impact histories where a collision of sufficient mass to produce prebiotic molecules does not occur, or alternatively, does occur but primitive life is subsequently destroyed by an ocean-vaporizing impact that does not rekindle the biosphere. A life-starting impact occurs 92\% or 41\% of the time when we assume a $> 4 \times 10^{20}$ kg or $> 5 \times 10^{21}$ kg impact without later ocean-vaporization is required to start life, respectively.

% Ocean vaporization caveat Caveat.

If life began in an RNA process after an impact, then this work suggests that origin of life on Earth was not an anomalous event and was indeed strongly favored (92\% chance) under optimistic assumptions. Given that rocky exoplanets accreted from impacts, our work supports and optimistic outlook for life on exoplanets and future searches for biosignatures.

@@

Overall, uncertainty in the impact properties required to vaporize the ocean has a small effect on our qualitative conclusions. Regardless of $f_\mathrm{E,vap}$, the origin of life on Earth from an impact does not appear to be anomalously improbable (Figure \ref{fig:probabilities_of_impacts_sens}) and biopoesis is preferred early in the Hadean with uncertainty spanning the entire eon (Figure \ref{fig:timing_sensitivity}). However, it is conceivable that a more complete understanding $f_\mathrm{E,vap}$ as a function of impact angle and other impact parameters could change these results.

>>>>

As illustrated in Figure \ref{fig:example_impact_histories}, a life-starting impact does not occur in every simulated impact history. In our impact-driven model for the origin of life, biopoesis requires (1) an impact of sufficient mass to make prebiotic nitriles and (2) the lack of a later ocean-vaporizing collision that destroys the biosphere without rekindling it. Figure \ref{fig:probabilities_of_impacts} shows the probability of conditions (1) and (2) occurring as a function of the minimum impact mass that produces prebiotic molecules. For example, consider the \citet{Wogan_2023} optimistic minimum mass of $4 \times 10^{20}$ kg indicated in Figure \ref{fig:probabilities_of_impacts}. All simulated impact histories have at least one impact bigger than $4 \times 10^{20}$ kg (blue line in Figure \ref{fig:probabilities_of_impacts}). However, the probability of the origin of life in this scenario is 92\%, because 8\% of the time the last $4 \times 10^{20}$ kg collision is followed by a smaller ocean-vaporizing impact that perhaps sterilizes the planet (orange dashed line in Figure \ref{fig:probabilities_of_impacts}). An origin of life is less probable if we consider the \citet{Wogan_2023} pessimistic minimum mass ($> 5 \times 10^{21}$ kg) to produce significant origin of life nitriles. A $> 5 \times 10^{21}$ kg impact without later ocean vaporization occurs 41\% of the time. A life-starting impact was therefore relatively likely on the early Earth, and was strongly favored (92\% probability) under optimistic assumptions.

@@

@@

Our simulations of Earth's impact history do not always result in an origin of life. There are some impact histories where a collision of sufficient mass to produce prebiotic molecules does not occur, or alternatively, does occur but primitive life is subsequently destroyed by an ocean-vaporizing impact that does not rekindle the biosphere. With our nominal assumptions, a life-starting impact occurs 92\% or 41\% of the time when we assume a $> 4 \times 10^{20}$ kg or $> 5 \times 10^{21}$ kg impact without later ocean-vaporization is required to start life, respectively. In our worst-case scenario, which assumes that impacts can more easily vaporize the ocean, a $> 5 \times 10^{21}$ kg impact without subsequent planet sterilization occurs in 20\% of simulated impact histories (Appendix Figure \ref{fig:probabilities_of_impacts_sens}).

We emphasize that these results rely upon a variety of assumptions including the premise that some of the first life-like molecules were RNA replicators which were built abiotically from nitriles like HCN and HCCCN. Additionally, our calculations assume that an impact that vaporizes the ocean also sterilizes the planet, but this may not be the case because microbes could have possibly survived in the deep subsurface \citep{Sleep_1989,Grimm_2018}.

If life began in an RNA-first process after an impact, then this work suggests that origin of life on Earth was a fairly likely event (at least a 20\% chance) and was indeed strongly favored (92\% chance) under optimistic assumptions. Given that all rocky planets form from accretionary impacts, our work supports and optimistic outlook for life on exoplanets and future searches for biosignatures.

@@

Overall, uncertainty in the impact properties required to vaporize the ocean has a small effect on our qualitative conclusions. Regardless of $f_\mathrm{E,vap}$, the origin of life on Earth from an impact does not appear to be a fluke (Figure \ref{fig:probabilities_of_impacts_sens}) and biopoesis is preferred early in the Hadean with uncertainty spanning the entire eon (Figure \ref{fig:timing_sensitivity}). However, it is conceivable that a more complete understanding $f_\mathrm{E,vap}$ as a function of impact angle and other impact parameters could change these results.
